Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-06-09 talk/icooolps2009/paper.tex
starting to incorporate reviewer comments

We attempt to apply the technique of Tracing JIT Compilers
in the context
of the PyPy project, \ie to programs that are interpreters for some
dynamic languages, including Python.  Tracing JIT compilers can greatly
speed up programs that spend most of their time in loops in which they
take similar code paths.  However, applying an unmodified tracing JIT to
a program that is itself a bytecode interpreter results in very limited
or no speedup.
In this paper we show how to guide tracing JIT compilers to greatly
improve the speed of bytecode interpreters.  One crucial point is to
unroll the bytecode dispatch loop, based on two hints provided by the
implementer of the bytecode interpreter.  We evaluate our technique by
applying it to two PyPy interpreters: one is a small example, and the
other one is the full Python interpreter.

@@

The PyPy project\footnote{http://codespeak.net/pypy}
\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} is an
environment where flexible implementation of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython \cite{ancona_rpython:step_2007}. RPython (``Restricted
Python'') is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLI and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

@@

An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator.  The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers.\footnote{The
chain of \texttt{if}, \texttt{elif}, ... instructions checking the various
opcodes is turned into a \texttt{switch} statement by one of PyPy's
optimizations. Python does not have a \texttt{switch} statement.}
A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

For the small example the hints look like a lot of work. However, the number of
hints is essentially constant no matter how large the interpreter is, which
makes the extra work negligible for larger interpreters.

@@

To test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-python}. While the function may seem a bit arbitrary, executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations, like indexing into the tuple, addition
and comparison of \texttt{i}. We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.2 (the reference implementation of
Python). In addition we benchmarked CPython using Psyco 1.6
\cite{rigo_representation-based_2004}, a specializing JIT compiler for Python.

>>>>

We attempt to apply the technique of Tracing JIT Compilers
in the context
of the PyPy project, \ie to programs that are interpreters for some
dynamic languages, including Python.  Tracing JIT compilers can greatly
speed up programs that spend most of their time in loops in which they
take similar code paths.  However, applying an unmodified tracing JIT to
a program that is itself a bytecode interpreter results in very limited
or no speedup.
In this paper we show how to guide tracing JIT compilers to greatly
improve the speed of bytecode interpreters.  One crucial point is to
unroll the bytecode dispatch loop, based on two kinds of hints provided by the
implementer of the bytecode interpreter.  We evaluate our technique by
applying it to two PyPy interpreters: one is a small example, and the
other one is the full Python interpreter.

@@

The PyPy project\footnote{http://codespeak.net/pypy}
\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} is an
environment where flexible implementations of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython \cite{ancona_rpython:step_2007}. RPython (``Restricted
Python'') is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLI and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

@@

An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator.  The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers.\footnote{The
chain of \texttt{if}, \texttt{elif}, ... instructions checking the various
opcodes is turned into a \texttt{switch} statement by one of PyPy's
optimizations. Python does not have a \texttt{switch} statement.}
A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
not improved at all.

@@

For the small example the hints look like a lot of work. However, the number of
hints that need to be put into the interpreter source remains small, which
makes the extra work negligible for larger interpreters.

@@

To test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-python}. While the function may seem a bit constructed, executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations, like indexing into the tuple, addition
and comparison of \texttt{i}. We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.2 (the reference implementation of
Python). In addition we benchmarked CPython using Psyco 1.6
\cite{rigo_representation-based_2004}, a specializing JIT compiler for Python.
